\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-18 paper.qmd
Add tSNE view of HARPS star data and videos with absolute error values for PBMC3k and HARPS data

### Tunning the model

@@

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/Ia3aL3L-fCk>)"
#| label: fig-nldervisPBMCUMAP
 #| out-width: 100%

@@

### Single-Cell Tagged Reverse Transcription sequencing data of mouse

@@

>>>>

### Tuning the model

@@

- learned from the model: There are 3 well-separated clusters in 2D, but if look at the model in high-D space, the three clusters are much closer. Also, there is some continuity within the clustering structure and it didn't capture well. Therefore UMAP with n:neighbour 30 is not a good representation for this data set.

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/lIA6ZPZAcLE>)"
#| label: fig-nldervisPBMCUMAP
 #| out-width: 100%

@@

### HARPS GTO sample dataset

- In 2D, there are two well-separated clusters

- One cluster have a sparse data space within the cluster

- There is a line in right corner of the structure that is not connected any other trimesh points. This line create a long edge in high-D space.

- In high-D view, the clusters having some non-linearity patter, which is also not captured in tSNE. Therefore, this is not a good choice.

tSNE_data_harps <- read_rds("data/harps/harps_tsne_40.rds")

plot_list2_harps <- tSNE_data_harps |>
    ggplot(aes(x = tSNE1,
               y = tSNE2, color = cluster))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#cab2d6", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  guides(color = guide_legend(title = "cluster"),
         color = guide_legend(nrow = 3)) #+
  # annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

## Import data
training_data_1 <- read_rds("data/harps/harps_data.rds")
training_data_1 <- training_data_1 |>
  mutate(ID = 1:NROW(training_data_1))

clusters <- training_data_1$cluster
select_names <-  names(training_data_1)[-c(18, length(training_data_1))]

training_data_1 <- training_data_1 |> select(-cluster)
names(training_data_1) <- append(paste0("x", 1:17), "ID")

tSNE_data_harps <- tSNE_data_harps |>
  select(tSNE1, tSNE2) |>
  mutate(ID = 1:NROW(tSNE_data_harps))

num_bins_tsne_harps <- calculate_effective_x_bins(.data = tSNE_data_harps, x = tSNE1,
                                                cell_area = 1) #43

shape_val_tsne_harps <- calculate_effective_shape_value(.data = tSNE_data_harps,
                                                      x = tSNE1, y = tSNE2)
## To extract bin centroids
hexbin_data_object_tsne_harps <- extract_hexbin_centroids(nldr_df = tSNE_data_harps, num_bins = num_bins_tsne_harps, shape_val = shape_val_tsne_harps, x = tSNE1, y = tSNE2)

df_bin_centroids_tsne_harps <- hexbin_data_object_tsne_harps$hexdf_data

## Identify bins with low-density
# identify_rm_bins <- find_low_density_hexagons(df_bin_centroids_tsne_harps, num_bins_tsne_harps, benchmark_rm_hex = 0.01)
#
# df_bin_centroids_tsne_harps <- df_bin_centroids_tsne_harps |>
#   filter(!(hexID %in% identify_rm_bins))

tSNE_data_with_hb_id_harps <- tSNE_data_harps |>
  dplyr::mutate(hb_id = hexbin_data_object_tsne_harps$hb_data@cID)

## To generate a data set with high-D and 2D training data
df_all_tsne_harps <- dplyr::bind_cols(training_data_1 |> dplyr::select(-ID), tSNE_data_with_hb_id_harps)

## Averaged on high-D
df_bin_tsne_harps <- avg_highD_data(.data = df_all_tsne_harps)

## Triangulate bin centroids
tr1_object_tsne_harps <- triangulate_bin_centroids(df_bin_centroids_tsne_harps, x, y)
tr_from_to_df_tsne_harps <- generate_edge_info(triangular_object = tr1_object_tsne_harps)

## Compute 2D distances
distance_tsne_harps <- cal_2D_dist(.data = tr_from_to_df_tsne_harps)

## To find the benchmark value
benchmark_tsne_harps <- find_benchmark_value(.data = distance_tsne_harps, distance_col = distance)

trimesh_removed_tsne_harps <- remove_long_edges(.data = distance_tsne_harps, benchmark_value = benchmark_tsne_harps,
                                              triangular_object = tr1_object_tsne_harps, distance_col = distance)

trimesh_removed_tsne_harps <- trimesh_removed_tsne_harps +
  # xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  # theme(axis.text = element_text(size = 5),
  #       axis.title = element_text(size = 7)) +
  geom_point(colour = "#33a02c", size = 0.5) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from tSNE (perplexity = 40) applied for the HARPS GTO dataset. Is this a best representation of the original data?, (b) Model in the 2D space with tSNE (<https://youtu.be/txTD7GTxt6Y>)"
#| label: fig-nldervisharpstSNE

plot_list2_harps + trimesh_removed_tsne_harps +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))

<!--
### Single-Cell Tagged Reverse Transcription sequencing data of mouse

@@

# 2D layouts from different NLDR techniques applied for the training PBMC dataset: (a) tSNE (perplexity = 30), (b) UMAP (n_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n_inliers = 5, n_outliers = 4, n_random = 3), and (e) PaCMAP (n_neighbors = 10, init = random, MN_ratio = 0.9, FP_ratio = 2). Is there a best representation of the original data or are they all providing  equivalent information?
```
-->

### 2024-01-18 paper.tex
Add tSNE view of HARPS star data and videos with absolute error values for PBMC3k and HARPS data

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{tunning-the-model}{%
\subsection{Tunning the model}\label{tunning-the-model}}

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\end{itemize}

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/Ia3aL3L-fCk})}

@@

\hypertarget{single-cell-tagged-reverse-transcription-sequencing-data-of-mouse}{%
\subsection{Single-Cell Tagged Reverse Transcription sequencing data of
mouse}\label{single-cell-tagged-reverse-transcription-sequencing-data-of-mouse}}

The Zeisel mouse brain dataset, obtained through Spatial Transcriptomics
(STRT-Seq). Within this dataset, information is collected from a
substantial 2,816 individual mouse brain cells. Each of these cells acts
as a molecular snapshot, capturing the distinctive genetic activity
within various cell types. This diversity spans neurons, glial cells,
and other essential components of the brain, offering a comprehensive
view of the cellular tapestry.

What makes this dataset particularly valuable is its ability to shed
light on the spatial distribution of cells. Researchers can explore how
gene expression patterns vary across different regions of the mouse
brain, unlocking insights into the functional specialization of these
regions and the intricate networks that underpin neural processes.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis5Mouse-1.pdf}

@@

\caption{\label{fig-nldervis5Mouse}2D layouts from different NLDR
techniques applied for the training Zeisel mouse brain dataset: (a) tSNE
(perplexity = 30), (b) UMAP (n\_neighbors = 15), (c) PHATE (knn = 5),
(d) TriMAP (n\_inliers = 5, n\_outliers = 4, n\_random = 3), and (e)
PaCMAP (n\_neighbors = 10, init = random, MN\_ratio = 0.9, FP\_ratio =
2). Is there a best representation of the original data or are they all
providing equivalent information?}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltZEI-1.pdf}

\caption{\label{fig-diagnosticpltZEI}Goodness of fit statistics from
different NLDR techniques applied to training Zeisel mouse brain
dataset. What is the best NLDR technique to represent the original data
in 2D?}

\hypertarget{retina-dataset}{%
\subsection{Retina dataset}\label{retina-dataset}}

Human retina:
https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-7316

https://www.cell.com/cms/10.1016/j.cell.2015.05.002/attachment/fd52950c-fd35-401a-9e60-ca7588ef9001/mmc1.pdf

https://www.cell.com/fulltext/S0092-8674(15)00549-8\#app3

32 PCs tSNE (perplexity 30)

\hypertarget{kang-dataset}{%
\subsection{Kang dataset}\label{kang-dataset}}

For the Kang lupus dataset (29), we obtained 16 cell types based on
clustering of the Festem- selected genes (Fig. 4E and Fig. S10, S11),
including 2 fine cell types that were often missed by other methods. The
2 cell types were IFNhi CD14+ monocytes with high expression of type I
interferon-related (IFN) genes, and HSP+ CD4 T cells with up-regulated
heat- shock-protein-related (HSP) genes. The HSP+ CD4 T cells might
result from cryopreservation and thawing of T cells (37). Other than
devianceFS, other available methods could not identify these HSP+ CD4 T
cells (Fig. S11).

{\centering \includegraphics{paper_files/figure-pdf/fig-nlderviskangtSNE-1.pdf}

\caption{\label{fig-nlderviskangtSNE}2D layout from tSNE (perplexity =
30) applied for the Kang dataset. Is this a best representation of the
original data?}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{tuning-the-model}{%
\subsection{Tuning the model}\label{tuning-the-model}}

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\end{itemize}

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/lIA6ZPZAcLE})}

@@

\hypertarget{harps-gto-sample-dataset}{%
\subsection{HARPS GTO sample dataset}\label{harps-gto-sample-dataset}}

\begin{itemize}
\item
  In 2D, there are two well-separated clusters
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
\end{itemize}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisharpstSNE-1.pdf}

@@

\caption{\label{fig-nldervisharpstSNE}(a) 2D layout from tSNE
(perplexity = 40) applied for the HARPS GTO dataset. Is this a best
representation of the original data?, (b) Model in the 2D space with
tSNE (\url{https://youtu.be/txTD7GTxt6Y})}
